Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance

### 2019-05-18 Manuscript/manuscript.Rmd
manuscript edits

We illustrate a unified approach toward preparing TAD boundary data for predictive modelling purposes from different cell lines and data resolutions (Figure 1). To enable us to build a model that suitably discriminates between regions on the linear genome that are near TAD boundaries and regions that are not, we created a cell line- and resolution-specific response variable that indicated whether a genomic bin of unit width overlapped with a flanked TAD boundary (Y=1), or not (Y=0). The unit width of each bin on the linear genome corresponded to the resolution of the contact matrix that domains were called on. Moreover, to account for uncertainty in boundary location, TAD boundaries were flanked on either side by the same unit length as the genomic bins.

To investigate the relationship between functional genomic annotations and the development of TAD boundaries, we devised 3 different predictor-types that composed the feature space of downstream random forest classification models. These predictor-type specific annotations represented the data matrix, **X** = {$X_{1}$, $X_{2}$, ..., $X_{p}$}, where $p$ is the complete set of functional annotations. Overlap counts was the least granular predictor-type considered, simply defined by the number of annotation-specific regions located within each genomic bin (Figure 2A). Overlap percents was a more refined predictor-type, defined as the total percent overlap between each annotation-specific region and genomic bins (Figure 2B). Lastly, distances, the most granular predictor-type, described the spatial relationship between the center of each genomic bin to the center of the nearest annotation-specific region (Figure 2C).  When measured in basepairs, distances were found to be highly skewed across all resolutions, for both the GM12878 and K562 cell lines (Supplementary Figures 1-2 A-D). As a normalization procedure, we performed a $log_{2}$ transformation, which resulted in much more gaussian-shaped distance densities (Supplementary Figures 1-2 E-H).

@@

[???probably better to report medians here]
For GM12878, the average number of genomic bins created across chromosomes was 12,214.1, 4,892.7, 2,450.9, and 1,227.5 for each resolution respectively (Table 1). Likewise, the average number chromosome specific TADs called on each resolution was found to be 375.0, 228.4, 115.1, and 50.6. These TADs resulted in an average of 1254.6, 714.0, 364.4, and 165.3 chromosome specific boundaries across each resolution. Each of these quantities was find to have a significantly large ICC, indicating that the number of genomic bins, TADs called, and corresponding unique boundaries were highly correlated between the different resolutions (Table 2). On the other hand, we found that the chromosome specific imbalance ratios remained consistenly severe across all resolution, with averages of 0.12, 0.17, 0.17, and 0.15 for each resolution respectively (ICC=0.114; p=0.109). Such low imbalance ratios indicated a wide disparity between the number of genomic bins that overlaped with a flanked TAD boundary and the much larger number that did not. The K562 cell line was found to exhibit similar profiles for the above quantities (Supplementary Tables 3-4) [??? Very bad and boring. You should tell the STORY, not the report. E.g., "Expectedly, the average number of genomic bins, TADs, and TAD boundaries increased with the increased resolution of Hi-C data (Table 1). However, the imbalance ratio, defined as ..., remained severe (~XXX) at all resolutions (Table 1). In summary, the problem of class imbalance exists at all resolutions of Hi-C data."]

@@

Using the Jaccard [??? Spell capital everywhere] index, we were able to determine which group-specific annotations were enriched together on the linear genome. As could be expected, for GM12878, the class of 14 annotations segmenting the chromation into different states exhibited very low pairwise correlations, with a jaccard index range of (0-0.001) (Table 1). [??? Wrong table?] However, specific sets of pairwise histone modifications and transcription factor binding sites (TFBS) were found to be highly enriched. The range of jaccard indices for the 10 histone modifications was found to be (0.022-0.539). Specifically, H3k27ac, H3k4me1, H3k4me2, and H3k4me3 were found to be highly correlated, high levels of which have been detected surrounding transcription start sites [@barski2007high]. For the 79 TFBS, the range of jaccard indices was (<0.001-0.425). Here, known mediators of loop extrusion components were found to have high pairwise correlations including CTCF, RAD21, and SMC3 [@sanborn2015chromatin; @fudenberg2016formation; @alipour2012self]. Similar profiles of jaccard indices were exhibited across the sets of annotations for the K562 cell line (Supplementary Table 4). <!--The high jaccard indices among select sets of genomic annotations gave precedent for the use of an elastic-net regularization algorithm in order to suitably reduce the feature space while retaining possibly correlated annotations that could be predictive of TAD boundary formation.-->

>>>>

We illustrate a unified approach toward preparing TAD boundary data for predictive modelling purposes from different cell lines and data resolutions (Figure 1). To enable us to build a model that suitably discriminates between regions on the linear genome that are near TAD boundaries and regions that are not, we created a cell line- and resolution-specific response variable that indicated whether a genomic bin of unit width overlapped with a flanked TAD boundary (Y=1), or not (Y=0). The unit width of each bin on the linear genome corresponded to the resolution of the contact matrix on which domains were called. Moreover, to account for uncertainty in boundary location, TAD boundaries were flanked on either side by the same unit length as the genomic bins.

To investigate the relationship between functional genomic annotations and the development of TAD boundaries, we devised 3 different predictor-types that composed the feature space of downstream random forest classification models. These predictor-type specific annotations represented the data matrix, **X** = {$X_{1}$, $X_{2}$, ..., $X_{p}$}, where $p$ is the complete set of functional annotations. Overlap counts was the least granular predictor-type considered, simply defined by the number of annotation-specific regions located within each genomic bin (Figure 2A). Overlap percents was a more refined predictor-type, defined as the total percent overlap between each annotation-specific region and genomic bins (Figure 2B). Lastly, distances, the most granular predictor-type, described the spatial relationship between the center of each genomic bin to the center of the nearest annotation-specific region (Figure 2C).  <!--When measured in basepairs, distances were found to be highly skewed across all resolutions, for both the GM12878 and K562 cell lines (Supplementary Figures 1-2 A-D).--> As a normalization procedure, we performed a $log_{2}$ transformation on the distances prior to modelling (See Methods Section).

@@

[???probably better to report medians here]
Expectedly, the average number of genomic bins, TADs called, and TAD boundaries increased with the increased resolution of cell-line specific Hi-C data (Table 1; Supplementary Tables 3). The correlation of each cell-line specific quantity with resolution was found to be significant, as evidenced by large intraclass correllation coefficients (ICC) (Table 2; Supplementary Tables 4). However, the class imbalance problem was found to be much more consistant across resolutions in both cell lines. Likewise, this relationship was not significantly correlated with resoltuion for either cell line, indicating that there exists a severe disparity between classes at all resolutions of Hi-C data.

[??? Very bad and boring. You should tell the STORY, not the report. E.g., "Expectedly, the average number of genomic bins, TADs, and TAD boundaries increased with the increased resolution of Hi-C data (Table 1). However, the imbalance ratio, defined as ..., remained severe (~XXX) at all resolutions (Table 1). In summary, the problem of class imbalance exists at all resolutions of Hi-C data."; !!! editted]

@@

Using the Jaccard index, we were able to determine which group-specific annotations were enriched together on the linear genome. As could be expected, for GM12878, the class of 14 annotations segmenting the chromation into different states exhibited very low pairwise correlations, with a Jaccard index range of (0-0.001) (Table 3). However, specific sets of pairwise histone modifications and transcription factor binding sites (TFBS) were found to be highly enriched. The range of Jaccard indices for the 10 histone modifications was found to be (0.022-0.539). Specifically, H3k27ac, H3k4me1, H3k4me2, and H3k4me3 were found to be highly correlated, high levels of which have been detected surrounding transcription start sites [@barski2007high]. For the 79 TFBS, the range of Jaccard indices was (<0.001-0.425). Here, known mediators of loop extrusion components were found to have high pairwise correlations including CTCF, RAD21, and SMC3 [@sanborn2015chromatin; @fudenberg2016formation; @alipour2012self]. Similar profiles of Jaccard indices were exhibited across the sets of annotations for the K562 cell line (Supplementary Table 4). <!--The high Jaccard indices among select sets of genomic annotations gave precedent for the use of an elastic-net regularization algorithm in order to suitably reduce the feature space while retaining possibly correlated annotations that could be predictive of TAD boundary formation.-->
